A Realistic Guide to Making Data Available Alongside Code to Improve Reproducibility.

### 2020-08-19 new-paper.Rmd
tinker with stubs
remove "we recommend"s and provide concrete statements.

1 = Monash University, Department of Econometrics and Business Statistics, Melbourne, Australia

2 = Australian Centre of Excellence for Mathematical and Statistical Frontiers (ACEMS)

3 = Berkeley Institute for Data Science, University of California, Berkeley, USA

* = Both authors contributed equally to the work

Correspondence to:

Nicholas Tierney (nicholas.tierney@gmail.com) and Karthik Ram (k@karthik.io)

@@

To make data more reusable, we need to make it easier to generate metadata, standardise good practices within and across communities, and make reusable data sharing a recognized, normalised, part of research. Here we present a set of best practices to enable this change. These include documentation (\@ref(documentation-readmes)), licensing (\@ref(licences)), citation (\@ref(citation-how-you-want-your-data-to-be-cited)), code (\@ref(data-as-part-of-open-source-packages)), raw and analysis ready data (\@ref(raw-data-vs-analysis-read-data)). We also discuss considerations for sharing of data of various sizes \@ref(considerations-for-data-size).

# Documentation (READMEs) {.readme}

A README is often the first place users will look to learn more about a directory [@Prana2019]. The README is a concise text file containing guidance on how to read and interpret the directory - the who, what, when, where, and how (<https://betta.io/blog/2017/02/07/developer-experience-github-readmes/> and [@Prana2019]. For example, a README in a data repository might explain which directories contain data, tidying scripts, and also why data was collected. The README should be placed in the top level of the project, and there can be multiple README files in a project. In the context of datasets, READMEs are particularly useful when there are no reliable standards.

# Licences {.licenses}

A license for data clearly establishes the rules for its use, and whether it can be modified or shared. Without a license, or with a non-standard license, these rules are unclear, and lead to issues with use, attribution and citation. We recommend using a license from Creative Commons (<https://creativecommons.org/>), specifically, we recommend CCBY and CC0 as they are commonly used, and provide simple rules for attribution, warranty, and use. We recommend against non-standard licenses, such as those custom written by the authors, as some terms can render them unusable. Once a license is chosen, a LICENSE file should be provided, containing the entire license in the top level of the directory. The license type should be mentioned in the README, if used.

@@

# Metadata {.metadata}

There are three benefits to a dataset having metadata: (1) They provide context by describing variables to humans; (2) They avoid errors by describing and preserving variable types for machine readability; (3) They make data easier to find online by facilitating indexing [@google-data-search]. Three pieces of metadata should be considered: (1) **variable names**, (2) **variable descriptions**, and (3) **unit text**. These should be curated by a person. Metadata created should be clearly labelled in a folder or in files named METADATA.

**Variable names** are short, descriptive names with no spaces or special characters, such as "job\_position" or "date". **Variable description** describes the measured variables, e.g., "University Job Position" or "Date Started", and should also define *variable codes* and *missing data*. **Unit text** defines the type of variable recorded (e.g., string, date, numeric, integer), and avoids serious errors of misrepresented data (e.g., a column of a gene sequence being interpreted as a date). For more details on this we recommend reading [@Broman2017; @Ellis2017; @Arslan2019].

@@

EML is widely used in Ecology (for example [@Maerz2014]). JSON-LD provides advantages over EML, by having fields to facilitate data discovery with google dataset search [@EML-data]. We recommend JSON-LD over EML, as it is more human readable, shorter, and quicker and easier to read and write. For more information on tooling to assist generating metadata, see <https://knb.ecoinformatics.org/tools/morpho> and [@Arslan2019].

@@

# Raw data vs analysis read data {.rawdata}

Raw data is the first format of data provided before any cleaning has taken place (e.g., data in some binary or proprietary format). It should be placed in its own folder. Data used in analysis should be provided, especially if computationally intense to derive. It should be in its own folder, and to maximise interoperability, recorded in a plain text (e.g., csv, tsv), not a binary format (e.g., .rds, .sav, .xlsx). Ideally data should be in "Tidy Data" form, where variables are in columns, observations in rows, and each cell has only one value [@Wickham2014].

We don't have particular recommendations for non tabular of data, like binary images, but given the ubiquity of tables in papers, these recommendations will be broadly useful. For other data formats we recommend using a copy of the most robust common format and avoid the latest tools that may be more brittle and likely to change in the future. For example, for an image, a TIFF image in addition to a custom photoshop file.

# Data cleaning scripts {.scripts}

Code used to transform data from its raw format into analysis ready form should be kept in the same place as raw data (Section \@ref(fig:dir-layout)). Ideally, this transformation should involve only scripted languages (e.g, R, Python, Matlab, etc), but any non-scripted data preparation steps should be recorded in a plain text file.

# Citation: How you want your data to be cited {.cite}

Citation is important for provenance, and credit. Citing data clarifies the data version used (provenance), and provides authors a citation (credit). To ensure these are provided, data needs to have an accession number, such as a Digital Object Identifier (DOI). A DOI is a permanent, unique identifier for a digital object such as a paper, poster, or software, and is a prerequisite for citation. If a DOI or accession number is unavailable, a citation will be meaningless, as it cannot be tracked by any means. A file named CITATION should be placed in the directory at the top level. It should contain a DOI, and could be in `.bibtex` format - \@ref(fig:dir-layout). Avenues that provide DOIs are discussed in \@ref(avenues-to-share-data)].

# Avenues to share data {.avenues}

@@

As discussed in \@ref(citation-how-you-want-your-data-to-be-cited), we recommend data are archived using a platform that generates an accession number such as a DOI. For sharing data collected for research, the most relevant homes are the research data repositories **zenodo**, **dryad**, or **Open Science Framework** (OSF). In addition to minting DOIs, they also provide citation templates that can be used in \@ref(citation-how-you-want-your-data-to-be-cited).

@@

There are many alternatives to these three, for example, an astronomy project may host its data at the Sloan Digital Sky Survey (SDSS), and genetic data can be hosted at GenBank [@GenBank]. Importantly, these all provide some form of accession number. We recommend adhering to community standards when deciding where to publish data, as this will likely improve data reuse. Large data may also be shared via domain specific venues (e.g., in genomics or astronomy), where appropriate.

We recommend against depositing data in local file storage that does not provide DOI or accession numbers. For example, [@50Hectares] is hosted on a university server and accessed via FTP. It is not possible to effectively provide effective provenance and credit.

@@

Journals provide data papers, where they provide a formal publication explaining data that is archived. This is a familiar avenue for researchers to receive credit and citation. Writing a data paper can be very useful for researchers, however, data papers lack guidelines on how to structure data for effective reuse and sharing. We recommend partnering with [@dryad] to archive data and link it with research papers, as they provide a person to help curate your data, improving usability.

# Data as part of open source packages {.pkgs}

@@

Packaged data and archival on venues like Zenodo are not mutually exclusive. We recommend researchers first archive data in a long-term data repository (Section \@ref(avenues-to-share-data)), before including it in a package. This makes the data easily citable wherever data is referenced.

# Considerations for data size {.size}

@@

Small data has more options in how it is shared - you can bundle it with your analysis code (as in \@ref(raw-data-vs-analysis-read-data)), or in software \@ref(data-as-part-of-open-source-packages). For medium data, consider a venue like zenodo for archiving. Sharing medium data might require it to be broken into separate chunks for reading (e.g., `teaching-1.csv`, `teaching-2.csv`, etc.), and scripts provided to read it. Large data will require liaising with services from \@ref(avenues-to-share-data), or local research institute library to make special arrangements, to ensure there is some way to gain access, and also maintain a DOI.

>>>>

1 = Monash University, Department of Econometrics and Business Statistics, Melbourne, Australia\\
2 = Australian Centre of Excellence for Mathematical and Statistical Frontiers (ACEMS)\\
3 = Berkeley Institute for Data Science, University of California, Berkeley, USA\\
* = Both authors contributed equally to the work\\
Correspondence to:\\
Nicholas Tierney (nicholas.tierney@gmail.com) and Karthik Ram (k@karthik.io)

@@

To make data more reusable, we need to make it easier to generate metadata, standardise good practices within and across communities, and make reusable data sharing a recognized, normalised, part of research. Here we present a set of best practices to enable this change. These include documentation (Section \@ref(readme)), licensing (Section \@ref(licenses)), citation (Section \@ref(cite)), raw and analysis ready data (Section \@ref(rawdata)), code (Section \@ref(scripts)), and sharing in packages (Section \@ref(pkgs)). We also discuss considerations for sharing of data of various sizes (Section \@ref(size)).

# Documentation (READMEs) {#readme}

Use a README file with your data project. A README is often the first place users will look to learn more about a directory [@Prana2019]. The README is a concise text file containing guidance on how to read and interpret the directory - the who, what, when, where, and how (<https://betta.io/blog/2017/02/07/developer-experience-github-readmes/> and [@Prana2019]. For example, a README in a data repository might explain which directories contain data, tidying scripts, and also why data was collected. The README should be placed in the top level of the project, and there can be multiple README files in a project. In the context of datasets, READMEs are particularly useful when there are no reliable standards.

# Licences {#licenses}

Provide a license with your data. A license for data clearly establishes the rules for its use, and whether it can be modified or shared. Without a license, or with a non-standard license, these rules are unclear, and lead to issues with use, attribution and citation. Use a license from Creative Commons (<https://creativecommons.org/>), specifically, CCBY and CC0 as they are commonly used, and provide simple rules for attribution, warranty, and use. Do not use non-standard licenses, such as those custom written by the authors, as some terms can render them unusable. Once a license is chosen, a LICENSE file should be provided, containing the entire license in the top level of the directory. The license type should be mentioned in the README, if used.

@@

# Metadata {#metadata}

Use metadata with your data. There are three benefits to a dataset having metadata: (1) They provide context by describing variables to humans; (2) They avoid errors by describing and preserving variable types for machine readability; (3) They make data easier to find online by facilitating indexing [@google-data-search]. Three pieces of metadata should be considered: (1) **variable names**, (2) **variable descriptions**, and (3) **unit text**. These should be curated by a person. Metadata created should be clearly labelled in a folder or in files named METADATA.

**Variable names** are short, descriptive names with no spaces or special characters, such as "job\_position" or "date". **Variable description** describes the measured variables, e.g., "University Job Position" or "Date Started", and should also define *variable codes* and *missing data*. **Unit text** defines the type of variable recorded (e.g., string, date, numeric, integer), and avoids serious errors of misrepresented data (e.g., a column of a gene sequence being interpreted as a date). For more details, see [@Broman2017; @Ellis2017; @Arslan2019].

@@

EML is widely used in Ecology (for example [@Maerz2014]). JSON-LD provides advantages over EML, by having fields to facilitate data discovery with google dataset search [@EML-data]. Use JSON-LD over EML, as it is more human readable, shorter, and quicker and easier to read and write. For more information on tooling to assist generating metadata, see <https://knb.ecoinformatics.org/tools/morpho> and [@Arslan2019].

@@

# Raw data vs analysis ready data {#rawdata}

Make raw and analysis ready data available. Raw data is the first format of data provided before any cleaning has taken place (e.g., data in some binary or proprietary format). It should be placed in its own folder. Data used in analysis should be provided, especially if computationally intense to derive. It should be in its own folder, and to maximise interoperability, recorded in a plain text (e.g., csv, tsv), not a binary format (e.g., .rds, .sav, .xlsx). Ideally data should be in "Tidy Data" form, where variables are in columns, observations in rows, and each cell has only one value [@Wickham2014].

We don't have particular recommendations for non tabular of data, like binary images, but given the ubiquity of tables in papers, these recommendations will be broadly useful. For other data formats use a copy of the most robust common format and avoid the latest tools that may be more brittle and likely to change in the future. For example, for an image, a TIFF image in addition to a custom photoshop file.

# Data cleaning scripts {#scripts}

Share code used to transform data from its raw format into analysis ready form. This should be kept in the same place as raw data (Section \@ref(fig:dir-layout)). Ideally, this transformation should involve only scripted languages (e.g, R, Python, Matlab, etc), but any non-scripted data preparation steps should be recorded in a plain text file.

# Citation: How you want your data to be cited {#cite}

Describe how you want your data to be cited. Citation is important for provenance, and credit. Citing data clarifies the data version used (provenance), and provides authors a citation (credit). To ensure these are provided, data needs to have an accession number, such as a Digital Object Identifier (DOI). A DOI is a permanent, unique identifier for a digital object such as a paper, poster, or software, and is a prerequisite for citation. If a DOI or accession number is unavailable, a citation will be meaningless, as it cannot be tracked by any means. A file named CITATION should be placed in the directory at the top level. It should contain a DOI, and could be in `.bibtex` format (Figure \@ref(fig:dir-layout)). Avenues that provide DOIs are discussed in Section \@ref(avenues).

# Avenues to share data {#avenues}

@@

As discussed in \@ref(citation-how-you-want-your-data-to-be-cited),archive data using a platform generating an accession number such as a DOI. For sharing data collected for research, the most relevant homes are the research data repositories **zenodo**, **dryad**, or **Open Science Framework** (OSF). In addition to minting DOIs, they also provide citation templates that can be used in \@ref(citation-how-you-want-your-data-to-be-cited).

@@

There are many alternatives to these three, for example, an astronomy project may host its data at the Sloan Digital Sky Survey (SDSS), and genetic data can be hosted at GenBank [@GenBank]. Importantly, these all provide some form of accession number. Adhere to community standards when deciding where to publish data, as this will likely improve data reuse. Large data may also be shared via domain specific venues (e.g., in genomics or astronomy), where appropriate.

Do not deposit data in local file storage that does not provide DOI or accession numbers. For example, [@50Hectares] is hosted on a university server and accessed via FTP. It is not possible to effectively provide effective provenance and credit.

@@

Journals provide data papers, where they provide a formal publication explaining data that is archived. This is a familiar avenue for researchers to receive credit and citation. Writing a data paper can be very useful for researchers, however, data papers lack guidelines on how to structure data for effective reuse and sharing. Partner with Dryad [@dryad] to link archived data with research papers. Dryad will also provide a person to help curate your data, improving usability.

# Data as part of open source packages {#pkgs}

@@

Packaged data and archival on venues like Zenodo are not mutually exclusive. First, archive data in a long-term data repository (Section \@ref(avenues)), before including it in a package. This makes the data easily citable wherever data is referenced.

# Considerations for data size {#size}

@@

Small data has more options in how it is shared - you can bundle it with your analysis code (as in \@ref(rawdata)), or in software \@ref(data-as-part-of-open-source-packages). For medium data, consider a venue like zenodo for archiving. Sharing medium data might require it to be broken into separate chunks for reading (e.g., `teaching-1.csv`, `teaching-2.csv`, etc.), and scripts provided to read it. Large data will require liaising with services from \@ref(avenues), or local research institute library to make special arrangements, to ensure there is some way to gain access, and also maintain a DOI.
